---
title: "RAS84 signature sample classification of TCGA LUAD cohort : Project 002"
author: "philip.east@crick.ac.uk"
date: "12/02/2019"
output: 
  html_document:
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE,message=FALSE,warning=FALSE)
```

This script determines mutational correlates across the TCGA LUAD 5 RAS
activity groups.

```{r r_init}
library( affy )
library( limma )
library( reshape2 )
library( scales )
library( RColorBrewer )
library( ComplexHeatmap )
library( gplots )
library( gridExtra )
library( ggthemes )
library( Hmisc )
library( vcd )
library( UpSetR )
library( monocle )
library( pls )
library( DESeq2 )
library( circlize )
library( ggforce )
library( tidyverse )
library( openxlsx )
library( broom )
library( DT )
library( ggrepel )

my_grey <- "#707173"
my_red <- "#993333"
my_orange <- "#f6ad6e"
my_green <- "#7ab51d"
my_lightgreen <- "#adcf82"
my_purple <- "#bb90bd"
my_blue <- "#4066aa"
my_lightblue <- "#336699"

plot_formatter <- function() {
    theme_bw( ) +
        theme( panel.grid.major = element_blank( ),
               panel.grid.minor = element_blank( ),
               panel.border = element_blank( ),
              panel.background = element_blank( ),
              axis.line = element_line(color = "black"),
              axis.line.x = element_line(color="black", size = 0.5 ),
              axis.line.y = element_line(color="black", size = 0.5 ),
              text = element_text( size = 12 ) )
}

tumour <- "LUAD"
sig_n <- "RAS84"

## data resources
TCGA_LUAD_RAS84_CLASS_SE_FILE <- file.path( "data", "objects", "TCGA_LUAD_RAS84_class_se_t.rds" )
RAS_SIG_HOME <- file.path( "data", "resources", "signatures", "ras" )
RAS84_FEATURE_MAP_FILE <- file.path( "data", "resources", "RAS84_feature_map.xlsx" )

RAG_VARIANTS_FILE <- file.path( "results", "rag_variants.xlsx" )
SIGNATURE_CLUSTER_MUTATIONS_FILE <- file.path( "results", "signature_cluster_mutations.xlsx" )                                                
## figures
RAG_VARIANT_MOSAIC_PLOTS_FILE <- file.path( "figures", "mosaic_variant.pdf" )
RAS84_RAG_KRASMUT_BARPLOT_FILE <- file.path( "figures", "RAS84_RAG_KRASmut_barplots.pdf" )
KRASMUT_LOGLIK_FILE <- file.path( "figures", "SIG_RAG_KRASmut_mutation_loglik.pdf" )
RAS84_RAG_HEATMAP_FILE <- file.path( "figures", "RAS84_RAG_heatmap.pdf" )
RAS84_RAG_VARIANT_BARPLOT_FILE <- file.path( "figures", "RAS84_RAG_VARIANT_barplot.pdf" )
RAS84_RAG_KRASMUT_TYPE_BARPLOT_FILE <- file.path( "figures", "RAS84_RAG_KRASmut_type_barlot.pdf" )
```

# Load data

Load the SummarizedExperiment TCGA LUAD expression data from the TCGA
sample_classification analysis. Load the TCGA compatable filtered
signatures from the TCGA LUAD validate signatures analysis. Load TCGA somatic
mutation details.

```{r data_in}
se_t <- readRDS( TCGA_LUAD_RAS84_CLASS_SE_FILE )
krasSigNames <- sub( "\\.csv", "", dir( RAS_SIG_HOME ) )
signatures <- lapply( krasSigNames, function( sigFile ) {
    read.table( file = file.path( RAS_SIG_HOME, paste( sigFile, "csv", sep = "." ) ),
               sep = "\t", stringsAsFactors = FALSE )[, 1 ]
} )
names( signatures ) <- make.names( krasSigNames )
ras84_map <- read.xlsx( RAS84_FEATURE_MAP_FILE )
signatures$RAS84 <- ras84_map$TCGA_gene_symbol
```

# RAS pathway alterations

Here we look at RAS index distributions across RAS pathway mutation
status. We use the RAS pathway membership definition from the
oncogenic alterations TCGA Cell paper references above. This is to
determine if any oncogenic RAS pathway driver correlates with RAS84
signature activity.

```{r ras_pathway_alt,fig.width=8}
ras_alt_df <- colData( se_t )[, c( "rowname", "ras_index", rasAlts ) ] %>%
    as.data.frame( ) %>%
    tidyr::gather( alteration, present, -rowname, -ras_index )

ras_alt_prev <- ras_alt_df %>% 
    group_by( alteration ) %>% 
    summarize( prevalence = sum( present ) / n() ) %>%
    filter( prevalence > 0.05 )

dodge <- position_dodge( width = 0.4 )
plotdat <- ras_alt_df %>%
    filter( alteration %in% ras_alt_prev$alteration )
ras_alt_gg <- ggplot( plotdat, aes( x = alteration, y = ras_index, color = present ) ) +
    geom_violin( position = dodge ) +
    geom_boxplot( position = dodge, width = 0.1 ) +
    theme( axis.text.x = element_text( angle = 90, hjust = 1 ) )
print( ras_alt_gg )
```
```{r ras_pathway_alt_lm}
lmfit <- plotdat %>%
    nest( -alteration ) %>%
    mutate(
        fit = map( data, ~ lm( ras_index ~ present, data = .x ) ),
        tidied = map(fit, tidy) ) %>% 
    unnest( tidied ) %>%
    filter( term != "(Intercept)" ) %>%
    mutate( padj = p.adjust( p.value, method = "fdr" ) ) %>%
    filter( padj < 0.05 ) %>%
    arrange( p.value )
DT::datatable( lmfit )
```

# Chisq Test

Here we test TCGA LUAD somatic mutation distributions across our 5 RAS
activity groups using a chisq test. Alterations with a significant
non-random distribution across the RAS activity groups are shown in
the table and mosaic plots below.

```{r chisq_test}
mutation_dat <- colData( se_t ) %>%
    as.data.frame( ) %>%
    dplyr::select(
               barcode,
               RAS84,
               contains( "AMP", ignore.case = FALSE ),
               contains( "DEL", ignore.case = FALSE ),
               contains( "MUT", ignore.case = FALSE ),
               contains( "FUSION", ignore.case = FALSE ) ) %>%
    tidyr::gather( variants, value, -barcode, -RAS84 )

mutation_counts <- mutation_dat %>%
    group_by( variants ) %>%
    dplyr::summarize( total = sum( value ) ) %>%
    filter( total > 10 )

mutation_dat <- mutation_dat %>%
    filter( variants %in% mutation_counts$variants )

mutation_tab <- mutation_dat %>%
    group_by( RAS84, variants ) %>%
    dplyr::summarize( mutant = sum( value ),
                      wildtype = sum( !value ) ) %>%
    gather( genotype, count, -RAS84, -variants ) %>%
    unite( colheader, RAS84, genotype, sep = "." ) %>%
    spread( colheader, count )

mutation_res <- mutation_dat %>%
    group_by( variants ) %>%
    dplyr::summarize( pval = chisq.test( value, RAS84 )$p.value ) %>%
    mutate( fdr = p.adjust( pval, method = "fdr" ) )

mutations_decide <- mutation_res %>%
    filter( fdr < 0.05 ) %>%
    filter( variants != "DEL.CDKN2B" )

dbase <- mutation_tab %>%
    left_join( mutation_res, by = "variants" ) %>%
    arrange( fdr ) %>%
    as.data.frame( )

write.xlsx( dbase, file = RAG_VARIANTS_FILE )
DT::datatable( dbase )
```

## Mosaic plots {.tabset .tabset-fade}

```{r mosaic,results='asis',echo=FALSE}
for( i in mutations_decide$variants ) {
    tab <- mutation_dat %>%
        filter( variants == i ) %>%
        dplyr::select( RAS84, value ) %>%
        table( )
    colnames( tab ) <- c( "wildtype", "mutant" )
    cat( "### ", i, "\n" )
    print( mosaic( tab, shade = TRUE, legend = TRUE, main = i ) )
    cat( " \n\n" )
}
```
```{r mosaic_to_file}
for( i in mutations_decide$variants ) {
    tab <- mutation_dat %>%
        filter( variants == i ) %>%
        dplyr::select( RAS84, value ) %>%
        table( )
    colnames( tab ) <- c( "wildtype", "mutant" )
    cairo_pdf( file = file.path( adat$plot.path, sub( "variant", i, RAG_VARIANT_MOSAIC_PLOTS_FILE ) ) )
    print( mosaic( tab, shade = TRUE, legend = TRUE, main = i,
                  gp = shading_hsv,
                  gp_args = list( h = c(28/360, 203/360 ), s = c( 1, 0 ), v = c( 1, 0.5 ),
                                 interpolate = c( 2, 4 ) ) ) )
    dev.off( )
}
```

# Enriched mutants in other signatures

```{r other_sigs,eval=FALSE}
mutation_dat_l <- map( names( signatures ), function( sig_n ) { 
    mutation_dat <- colData( se_t ) %>%
        as.data.frame( ) %>%
        dplyr::select(
               barcode,
               sig_n,
               contains( "AMP", ignore.case = FALSE ),
               contains( "DEL", ignore.case = FALSE ),
               contains( "MUT", ignore.case = FALSE ),
               contains( "FUSION", ignore.case = FALSE ) ) %>%
        tidyr::gather( variants, value, -barcode, -sig_n ) %>%
        select( -barcode ) %>%
        rename( RGS = sig_n ) %>%
        filter( variants %in% mutations_decide$variants )
} ) %>%
    setNames( names( signatures ) )

chisq_res_l <- map( names( mutation_dat_l ), function( sig_n ) { 
    mutation_dat <-  mutation_dat_l[[ sig_n ]]
    mutation_counts <- mutation_dat %>%
        group_by( variants ) %>%
        dplyr::summarize( total = sum( value ) ) %>%
        filter( total > 10 )
    mutation_dat <- mutation_dat %>%
        filter( variants %in% mutation_counts$variants )
    chisq_res <- split( mutation_dat, mutation_dat$variants ) %>%
        map( function( x ) select( x, -variants ) ) %>%
        map( table ) %>%
        map( chisq.test ) %>%
        map( function( x ) x$p.value ) %>%
        unlist( )
    data.frame( signatures = sig_n,
               pval = chisq_res,
               fdr = p.adjust( chisq_res, method = "fdr" ),
               variant = names( chisq_res ) )
} )
names( chisq_res_l ) <- names( mutation_dat_l )

mutations_decide_all <- chisq_res_l %>%
    map( filter, fdr < 0.05 ) %>%
    bind_rows() %>%
    group_split( variant ) %>%
    map( arrange, pval )
names( mutations_decide_all ) <- mutations_decide_all %>%
    map( function( df ) df$variant[ 1 ] ) %>%
    unlist( )

gg <- bind_rows( chisq_res_l ) %>%
    ggplot( aes( x = variant, y = -log10( pval ), color = signatures, group = signatures ) ) +
    geom_point( ) +
    geom_line()
print( gg )

dbase_allsigs <- chisq_res_l %>%
    bind_rows( ) %>%
    write.xlsx( SIGNATURE_CLUSTER_MUTATIONS_FILE )
```
```{r other_sigs_bar}
gg_bar_df <- mutation_dat_l %>%
    bind_rows( .id = "signature" ) %>%
    filter( variants == "MUT.KRAS" ) %>%
    group_by( RGS, signature ) %>%
    summarize( mut_pc = sum( value )/n()*100 ) %>%
    mutate( signature = factor( signature ) ) 
gg_bar_df$signature <- relevel( gg_bar_df$signature, "RAS84" )

gg_bar <- ggplot( gg_bar_df, aes( x = RGS, y = mut_pc ) ) +
    geom_bar( stat = "identity", fill = my_green ) +
    facet_grid( signature ~ 1 ) +
    theme( strip.text.y = element_text( angle = 0 ),
           axis.text.x = element_text( angle = 90, hjust = 0.95 ) )
print( gg_bar )
```

```{r other_sigs_heatmap}
hmdat_l <- mutation_dat_l %>%
    map( function( df ) {
        df %>%
            group_by( RGS, variants ) %>%
            summarize( mutants = sum( value ),
                       N = n( ) ) %>%
            mutate( mutant_group_pc = mutants / N * 100 )
    } ) %>%
    bind_rows( .id = "signature" ) %>%
    split( .$variants ) %>%
    map( dplyr::select, -mutants, -N, -variants ) %>%
    map( spread, signature, mutant_group_pc )

hmcols <- colorpanel( 100, "white", "orange", "red" )

hm_l <- map( names( hmdat_l ), function( hmdat_n ) {
    hmdat <- hmdat_l[[ hmdat_n ]]
    hm_mat <- hmdat %>%
        column_to_rownames( var = "RGS" ) %>%
        as.matrix( )
    Heatmap( hm_mat,
            name = "percentage",
            column_title = hmdat_n,
            rect_gp = gpar(col = "grey" ),
            clustering_method_rows = "ward.D2",
            cluster_columns = FALSE,
            cluster_rows = FALSE,
            show_column_names = TRUE,
            show_row_names = TRUE,
            col = colorRamp2(
                seq( from = 0, to = 100, length.out = length( hmcols ) ),
                hmcols ) )
} )

hm <- hm_l[[ 1 ]]
for( i in 2:length( hm_l ) )
    hm <- hm + hm_l[[ i ]]
print( hm )
```

# KRAS mutation

KRAS mutation distribution across the 5 RAS activity groups for RAS84 segregation.

```{r kras_mutation}
plotdat <- colData( se_t )[, c( "rowname", "RAS84", "MUT.KRAS" ) ] %>%
  as.data.frame( ) %>%
  group_by( RAS84, MUT.KRAS ) %>%
  mutate( n = n( ) ) %>%
  group_by( RAS84 ) %>%
  mutate( N = n( ) ) %>%
  mutate( percent = n / N * 100 ) %>%
  dplyr::select( -rowname ) %>%
  distinct( )

gg <- ggplot( plotdat, aes( x = RAS84, y = percent, fill = MUT.KRAS, group = MUT.KRAS ) ) +
    geom_bar( stat = "identity" ) +
    scale_fill_manual( values = c( 'TRUE' = my_green, 'FALSE' = "white" ) ) +
    labs( x = "RAS Activity", y = "% oncogenic KRAS mutation" )

gg <- my_gg_themes( gg )
gg <- gg + theme( axis.text.x = element_text( angle = 90, hjust = 0.95 ) ) +
    theme( text = element_text( size = 14 ) )
gg

pdf( file = RAS84_RAG_KRASMUT_BARPLOT_FILE, height = 5, width = 2 )
print( gg )
dev.off( )
```

## KRAS mutation distribution (all signatures)

We quantify the effect size of the segragation of KRAS mutation across
the five RAG groups for all RAS signatures. We use log-likleyhood to
quantify segregation.

```{r kras_loglik}
KRAS_mutation_dat_l <- map( names( mutation_dat_l ), function( sig_n ) { 
    mutation_dat_l[[ sig_n ]] %>%
        filter( variants == "MUT.KRAS" )
} )
names( KRAS_mutation_dat_l ) <- names( mutation_dat_l )

glm_fit <- map( KRAS_mutation_dat_l, function( df ) {
    glm( value ~ RGS, data = df, family = "binomial" )
} )

plotdat <- map( glm_fit, logLik )%>%
    melt( ) %>%
    dplyr::rename( signatures = L1 )

loglik_gg <- ggplot( plotdat, aes( x = 1, y = value, label = signatures ) ) +
    geom_vline( xintercept = 1 ) +
    geom_point( size = 5 ) +
    geom_text_repel( nudge_x = 0.1 ) +
    plot_formatter( ) +
    theme( legend.position = "none",
          text = element_text( size = 8 ) )

cairo_pdf( file = sub( "RAS84", i, KRASMUT_LOGLIK_FILE ), width = 3, height = 3 )
print( loglik_gg )
dev.off( )
print( loglik_gg )
```

## Other mutations

```{r other_loglik}
loglik_other_gg_l <- map( names( mutations_decide_all ), function( mutation_gene ) {
    specific_mutation_dat_l <- map( names( mutation_dat_l ), function( sig_n ) { 
        mutation_dat_l[[ sig_n ]] %>%
            filter( variants == mutation_gene )
    } )
    names( specific_mutation_dat_l ) <- names( mutation_dat_l )
    glm_fit <- map( specific_mutation_dat_l, function( df ) {
        glm( value ~ RGS, data = df, family = "binomial" )
    } )
    plotdat <- map( glm_fit, logLik )%>%
        melt( ) %>%
        dplyr::rename( signatures = L1 )
    ggplot( plotdat, aes( x = 1, y = value, label = signatures ) ) +
        geom_vline( xintercept = 1 ) +
        geom_point( size = 5 ) +
        geom_text_repel( nudge_x = 0.1 ) +
        plot_formatter( ) +
        theme( legend.position = "none",
              text = element_text( size = 8 ) )
} )
names( loglik_other_gg_l ) <- names( mutations_decide_all )
```
```{r render_other_loglik}
for( i in names( loglik_other_gg_l ) ) {
    cairo_pdf( file = sub( "SIG", i, KRASMUT_LOGLIK_FILE ), width = 3, height = 3 )
    print( loglik_other_gg_l[[ i ]] )
    dev.off( )
}
```

## RAS pathway alteration burden

Here we look at the degree to which the RAS pathway as a whole is alterated per-patient and if this correlates with the RAS activity groupings. From the mosaic plot below we see that there is not a strong association of RAS pathway alteration burden and RAS activity groups.

```{r ras_pathway}
ras_pathway_n <- colData( se_t ) %>%
    as.data.frame( ) %>%
    dplyr::select( c( rasAlts, "RAS84" ) ) %>%
    rownames_to_column( var = "barcode" ) %>%
    gather( alteration, presence, -barcode, -RAS84 ) %>%
    group_by( barcode ) %>%
    summarize( count = sum( presence ) ) %>%
    ungroup( ) %>%
    arrange( count ) %>%
    left_join( as.data.frame( colData( se_t )[, c( "barcode", "RAS84" ) ] ),
              by = "barcode" )

mosaic( table( ras_pathway_n[, c( "count", "RAS84" ) ] ),
        shade = TRUE, legend = TRUE )
```

# Heatmap TCGA LUAD RAS activity 5 clusters 

Here we see other significant mutations associated with RAS activity
groups. First alongside the patient groups heatmap and then as barcharts.

```{r heatmap_dat}
hmdat <- assays( se_t )$vst[ rowData( se_t )$RAS84, ] %>%
                      as.data.frame( ) %>%
                      rownames_to_column( var = "gene_id" ) %>%
                      mutate( gene_id = sub( "\\|.*", "", gene_id ) )

hmmat <- hmdat %>%
    column_to_rownames( "gene_id" ) %>%
    as.matrix( )

hmmat <- hmmat - rowMedians( hmmat )
```

## Sample mutation annotation

Here we setup the significant mutations as row annotations for the heatmap. 

```{r mut_anno}
mutations <- mutations_decide %>%
    pull( variants )
mutations <- mutations[ c( 6, 4, 8, 5, 7, 9, 1, 2, 3 ) ]

anno_rows_muts_df <- colData( se_t )[, mutations ] %>%
    as.data.frame( )

cols_map <- sapply( colnames( anno_rows_muts_df ), function( x ) {
    if( x == "MUT.KRAS" ) {
        c( 'TRUE' = my_purple, 'FALSE' = "white")
    } else {
        c( 'TRUE' = my_green, 'FALSE' = "white")
    }
}, simplify = FALSE )

hm_anno_row <- rowAnnotation( df = anno_rows_muts_df,
                             col = cols_map,
                             na_col = "white",
                             gap = unit( 1, "mm" ),
                             show_legend = FALSE,
                             show_annotation_name = TRUE,
                             annotation_name_side = "top",
                             annotation_name_gp = gpar( fontsize = 6 ),
                             simple_anno_size = unit( 2, "mm" ),
                             border = TRUE )

hmcols <- colorpanel( 1000, "white", my_grey, "#E3001A" )
hm_ann_row_2_dat <- ras_pathway_n %>%
    dplyr::select( barcode, count ) %>%
    column_to_rownames( var = "barcode" ) %>%
    as.matrix( )
```

## Signature gene annotation

Here we setup parent signature gene membership as column annotation
for the heatmap.

```{r gene_anno}
RAS84_parent_signatures <- c( "HRAS", "KRASG13D134", "RAS_pathway" )

parents <- melt( signatures[ RAS84_parent_signatures ], value.name = "gene_id" ) %>%
    filter( gene_symbol %in% signatures$RAS84 ) %>%
    mutate( flag = TRUE ) %>%
    spread( L1, flag, fill = FALSE )

hmdat <- hmdat %>%
    left_join( parents, by = "gene_id" )

col_anno_dat <- dplyr::select_( hmdat, .dots = RAS84_parent_signatures )
colnames( col_anno_dat ) <- sig_name_map[ colnames( col_anno_dat ) ]
col_anno_colours <- lapply( colnames( col_anno_dat ),
                           function( x ) c( 'TRUE' = "lightgrey", 'FALSE' = "white" ) )
names( col_anno_colours ) <- colnames( col_anno_dat )

hm_anno_col <- HeatmapAnnotation( df = col_anno_dat,
                                 show_annotation_name = TRUE,
                                 simple_anno_size = unit( 1, "mm" ),
                                 col = col_anno_colours,
                                 annotation_width = 0.01,
                                 na_col = "white",
                                 gap = unit( 1, "mm" ),
                                 show_legend = FALSE,
                                 annotation_name_gp = gpar( fontsize = 6 ),
                                 annotation_name_side = "left" )
```

## Heatmap

```{r render_heatmap}
hmcols <- colorpanel( 1000, my_blue, "white", my_red )

hm <- Heatmap( t( hmmat ),
              name = "Expression", 
              clustering_method_rows = "ward.D2",
              clustering_method_columns = "ward.D2",
              show_column_names = TRUE,
              show_row_names = FALSE,
              col = colorRamp2(
                  seq( from = -2, to = 2, length.out = length( hmcols ) ),
                  hmcols ),
              row_split = se_t$RAS84,
              row_gap = unit( 1, "mm" ),
              column_title = "RAS84",
              column_names_gp = gpar( fontsize = 4 ),
              heatmap_legend_param = list( title_gp = gpar( fontsize = 14 ),
                                           labels_gp = gpar( fontsize = 4 ) ),
              show_heatmap_legend = TRUE,
              bottom_annotation = hm_anno_col,
              cluster_row_slices = FALSE,
              row_names_max_width = unit( 10, "mm" ) )

hmmat_ras_index <- matrix( se_t$ras_index - median( se_t$ras_index ), ncol = 1 )
rownames( hmmat_ras_index ) <- se_t$barcode

hm_ras_index <- Heatmap( hmmat_ras_index,
                        name = "RSI",
                        clustering_method_rows = FALSE,
                        clustering_method_columns = FALSE,
                        show_column_names = TRUE,
                        show_row_names = FALSE,
                        col = colorRamp2(
                            seq( from = -1, to = 1, length.out = length( hmcols ) ),
                            hmcols ),
                        row_split = se_t$RAS84,
                        row_gap = unit( 1, "mm" ),
                        column_names_gp = gpar( fontsize = 4 ),
                        show_heatmap_legend = TRUE )

print( hm + hm_ras_index + hm_anno_row )

pdf( file = RAS84_RAG_HEATMAP_FILE, width = 8, height = 4 )
hm + hm_ras_index + hm_anno_row
dev.off( )
```

# Significant mutation barcharts {.tabset .tabset-fade}

```{r sig_mut_barchart}
plotdat <- mutation_dat %>%
    filter( variants %in% mutations_decide$variants ) %>%
    group_by( RAS84, variants ) %>%
    summarize( mutant_percent = sum( value ) / n() * 100 )

sig_mut_gg_l  <- map( unique( plotdat$variants ), function( i ) {
    ggplot( subset( plotdat, variants == i ),
                 aes( x = RAS84 y = mutant_percent ) ) +
        geom_bar( stat = "identity", color = my_green, fill = my_green,
                  width = 0.8 ) +
        labs( x = "RAS Activity", y = paste( "% oncogenic", i, "mutation" ) ) +
        plot_formatter( ) +
        theme( axis.text.x = element_text( angle = 90, hjust = 0.95 ) )
} )
names( sig_mut_gg_l ) <- unique( plotdat$variants )
```
```{r sig_mut_barchart_out,results='asis',echo=FALSE}
for( i in names( sig_mut_gg_l ) ) {
    cat( "## ", i, "\n" )
    print( sig_mut_gg_l[[ i ]] )
    cat( " \n\n" )
}
```
```{r sig_mut_barchart_file}
for( i in names( sig_mut_gg_l ) ) {
    pdf( file = sub( "VARIANT", i, RAS84_RAG_VARIANT_BARPLOT_FILE ), height = 4, width = 1.5 )
    print( gg )
    dev.off( )
}
```

# KRAS specific variant distributions 

Here we look at the distribution of specific KRAS mutation types
across the RSGs. We remove RSG-0 and focus on the groups with
significant numbers of KRAS mutations. We run chisq.test globally and
then on the individual genotypes compared to the remaining samples,
treated as the control group.

```{r}
kras_onco_muts <- c( "p.G12A", "p.G12C", "p.G12D",
                    "p.G12R", "p.G12S", "p.G12V",
                    "p.G13C", "p.G13D", "p.Q61H", "p.Q61L" )

chisq_dat <- data.frame( RAS84 = se_t$RAS84,
                        amino = as.character( se_t$amino ),
                        stringsAsFactors = FALSE ) %>%
    filter( RAS84 != "RSG_0" ) %>%
    droplevels( )

chisq_dat$amino[ !chisq_dat$amino %in% kras_onco_muts ] <- NA
chisq_dat$amino[ is.na( chisq_dat$amino ) ] <- "wt"

print( mosaic( table( chisq_dat ), shade = TRUE, legend = TRUE, main = " KRAS variant ~ RAS activity groups" ) )
pdf( file = file.path( adat$plot.path, "KRAS_variant_genotype_mosaic.pdf" ) )
mosaic( table( chisq_dat ), shade = TRUE, legend = TRUE, main = " KRAS variant ~ RAS activity groups" )
dev.off( )

chisq.test( table( chisq_dat ) )

chisq_res <- map( kras_onco_muts, function( mut ) {
    chisq_dat_local <- chisq_dat %>%
        mutate( amino = ifelse( amino == mut, amino, "wt" ) )
    chisq.test( table( chisq_dat_local ) )
} )
names( chisq_res ) <- kras_onco_muts

map( chisq_res, function( x ) data.frame( pval = x$p.value ) ) %>%
    bind_rows( .id = "KRAS_genotypes" ) %>%
    datatable( rownames = FALSE )

chisq_res <- map( chisq_tab, chisq.test )

ras_group_N <- colData( se_t ) %>%
    as.data.frame( ) %>%
    group_by( RAS84 ) %>%
    summarize( N = n( ) )

plotdat <- colData( se_t )[, c( "RAS84", "amino" ) ] %>%
    as.data.frame( ) %>%
    filter( amino %in% c( "p.G12A", "p.G12C", "p.G12D",
                         "p.G12R", "p.G12S", "p.G12V",
                         "p.G13C", "p.G13D", "p.Q61H",  "p.Q61L" ) ) %>%
    droplevels( ) %>%
    group_by( RAS84 , amino ) %>%
    summarize( n = n( ) ) %>%
    left_join( ras_group_N, by = "RAS84" ) %>%
    mutate( percentage = n / N * 100 )

gg <- ggplot( plotdat, aes( x = RAS84, y = percentage, group = amino, fill = amino ) ) +
    geom_bar( stat = "identity" ,color = "grey" ) +
  labs( x = "RAS activity", y = "% KRAS mutation type", fill = "KRAS mutation" ) +
  plot_formatter( )

gg <- gg + theme( axis.text.x = element_text( angle = 90, hjust = 0.95 ) ) +
    theme( text = element_text( size = 12 ) ) +
    theme( legend.position = "right" )
print( gg )

pdf( file = RAS84_RAG_KRASMUT_TYPE_BARPLOT_FILE, height = 5, width = 3 )
print( gg )
dev.off( )
```
