---
title: "Pancancer : Project 002"
author: "philip.east@crick.ac.uk"
date: 'Compiled: `r format(Sys.time(), "%d %B, %Y @ %H:%M:%S")`'
output:
  html_document:
    df_print: tibble
    toc: true
    toc_depth: 5
    toc_float: true
    code_folding: hide
    number_sections: yes
---

```{css setup_css, echo=FALSE}
body .main-container {
  max-width: 80%;
}
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

Presented here is a pan-TCGA RAS activity analysis. First we calculate
a RAS index value for each tumour (RAS84 signature mean) across a
pan-TCGA normalised expression data set. Second we classify
tumours, from each TCGA tumour cohort, into three RAS activity groups
(RASsig_max, RASsig_1, and RASsig_0). The RAS index distributions for
each tumour and each RAS activity group are shown below. We then
explore the link between RAS activity and overall survival for each
tumour type. The hypothesis being that RAS activty, in the more RAS
driven tumours (where the signature is likely to be more effective), is more
likely to correlate with overall survival. 

```{r r_init,message=FALSE,warning=FALSE}
library( affy )
library( limma )
library( org.Hs.eg.db )
library( ggplot2 )
library( reshape2 )
library( scales )
library( RColorBrewer )
library( ComplexHeatmap )
library( circlize )
library( colorspace )
library( GetoptLong )
library( dendextend )
library( matrixStats )
library( DESeq2 )
library( gplots )
library( ggplot2 )
library( gridExtra )
library( tibble )
library( dplyr )
library( tidyr )
library( gtable )
library( openxlsx )
library( survival )
library( survminer )
library( ggrepel )
library( DT )
library( broom )
library( purrr )

source( "/camp/stp/babs/working/eastp/code/R/R.packages.east01.functions.R")
source( "../../scripts/lib.Ras_HighLow.335.R" )

adat <- project.init( "results", root = "." )

my_grey <- "#707173"
my_red <- "#cc3333"
my_orange <- "#ff9933"

my_lightgreen <- "#adcf82"
my_purple <- "#bb90bd"
my_blue <- "#6699cc"

plot_formatter <- function() {
    theme_bw( ) +
        theme( panel.grid.major = element_blank( ),
               panel.grid.minor = element_blank( ),
               panel.border = element_blank( ),
               panel.background = element_blank( ),
               #axis.line = element_line(color = "black"),
               #axis.line.x = element_line(color="black", size = 0.1 ),
               #axis.line.y = element_line(color="black", size = 0.1 ),
               text = element_text( size = 10 ) )
}
```

# Load metadata

Load the RAS84 signature and the TCGA Pan-Cancer Oncogenic pathway
alteration data and survival data from TCGA Pan-Cancer Immune
Landscape paper. We also read in the curated survival data from
[TCGA-CRD](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC6066282/)
project where they derive PFI values thay they show are more
appropriate across more tumour types than OS.

```{r project_init,message=FALSE,warning=FALSE}
SIG_HOME <- file.path( "data", "signatures" )
krasSigNames <- sub( "\\.csv", "", dir( SIG_HOME ) )
signatures <- lapply( krasSigNames, function( sigFile ) {
    read.table( file = file.path( SIG_HOME, paste( sigFile, "csv", sep = "." ) ),
               sep = "\t", stringsAsFactors = FALSE )[, 1 ]
} ) %>% setNames( make.names( krasSigNames ) )
RAS84_signature_map <- read.xlsx( file.path( "data", "RAS84_feature_map.xlsx" ) )
signatures$RAS4 <- RAS84_signature_map$TCGA_gene_symbol

## Read in pan-cancer mutation data
alt_altDatFile <- file.path( "data", "oncogenicRASPath_TCGA", "oncogenic_pathway_alteration_level_TCGA.txt" )
alt_altDat <- read.table( file = alt_altDatFile, header = TRUE, sep = "\t" )
colnames( alt_altDat )[ 1 ] <- "pancan_pathway"

rasPathwayDefFile <- file.path( "data", "rasPathwayDefinition.txt" )
rasPathwayTab <- read.delim( file = rasPathwayDefFile, sep = "\t" )

rasGeneSets <- list( og = subset( rasPathwayTab, OG.TSG == "OG" )$Gene,
                    tsg = subset( rasPathwayTab, OG.TSG == "TSG" )$Gene )
rasGeneSets <- lapply( rasGeneSets, as.character )
ras_pathway_genes <- unlist( rasGeneSets )

## identify RAS pathway alterations
alt_genes <- sapply( strsplit( colnames( alt_altDat )[ -1 ], "\\." ), '[[', 2 )
rasAlts_f <- alt_genes %in% ras_pathway_genes
rasAlts <- colnames( alt_altDat )[ -1 ][ rasAlts_f ]

tcga_crd_dat <- read.xlsx( file.path( "data", "TCGA-pancancer_CDR_Lui_et_al_supptab1.xlsx" ),
                           sheet = "ExtraEndpoints" )
```

# Read tumour expression data

```{r read_tumour_data}
se_l <- readRDS( file.path( "data", "se_l.rda" ) )    load( file = se_l_file )
## remember to - assays( se_t )$zscore <- apply( assays( se_t )$vst, 2, function( x ) ( x - median( x ) )/ mad( x ) )

se_pc <- readRDS( file.path( "data", "se_pancancer.rds" ) )
```

Here we calculate the mean RAS index for each tumour cohort. This
will be used to order the tumours throughout this analysis. These
values are shown in the barchart below.

```{r ras_index,message=FALSE,warning=FALSE}
sig_f <- rowData( se_pc )$gene_id %in% signatures$RAS84
se_pc$ras_index_pc <- colMeans( assays( se_pc )$zscore[ sig_f, ] )
```

Shown here are the per-tumour cohort RAS index distributions. The
first is the original plot where each point represents the RAS index
value for a particular tumour and the red indicates an oncogenic RAS
mutant (K,H,N). The second plot shows the RAS index distributions for
each of the RAS activity groups (RASsigmax=red, RASsig_1=green,
RASsig_0=blue). RAS mutants are highlighted by the black dots.

## Per cohort RAS summary statistics

Here we calculate RAS mutation percentages and mean RAS index values
per cohort.

```{r ras_index_data,message=FALSE,warning=FALSE,fig.height=8}
plotdat_violin <- colData( se_pc ) %>%
  as.data.frame( ) %>%
    dplyr::select( barcode, tumour, ras_index_pc, rasact_3,
                   MUT.KRAS, MUT.NRAS, MUT.HRAS, AMP.HRAS, AMP.KRAS ) %>%
  mutate( mut_ras_type = ifelse( MUT.KRAS, "kras",
                         ifelse( MUT.NRAS, "nras",
                         ifelse( MUT.HRAS, "hras", "wt" ) ) ) ) %>%
  mutate( MUT.RAS = mut_ras_type != "wt" ) %>%
  mutate( AMP.RAS = AMP.HRAS | AMP.KRAS ) %>%
  mutate( mut_ras_type = ifelse( MUT.RAS, mut_ras_type,
                                ifelse( AMP.RAS, "amp", "wt" ) ) )
plotdat_violin$MUT.RAS[ is.na( plotdat_violin$MUT.RAS ) ] <- FALSE

ras_sumstats_dat <-  plotdat_violin %>%
    group_by( tumour ) %>%
    summarize( RASmut = ( sum( mut_ras_type  != "wt" )/n() ) * 100,
               mean_ras_index = mean( ras_index_pc ),
               median_ras_index = median( ras_index_pc ),
              cases = n() ) %>%
    ungroup( ) %>%
    mutate( high_ras_mut = as.factor( RASmut > 7.5 ) ) %>%
    arrange( mean_ras_index ) %>%
    mutate( tumour = factor( tumour, levels = rev( tumour ) ) ) %>%
    mutate( cuts = cut( mean_ras_index, 4,
                       c( "q_lower", "q_2", "q_3", "q_upper" ),
                       include.lowest = TRUE ) ) %>%
    arrange( mean_ras_index ) %>%
    mutate( tumour = factor( tumour, levels = tumour ) ) %>%
    mutate( quartiles = cut( mean_ras_index,
                            c( min( mean_ras_index ),
                              quantile( mean_ras_index, c( 0.25, 0.5, 0.75 ) ),
                              max( mean_ras_index ) ),
                            c( "q_lower", "q_2", "q_3", "q_upper" ),
                            include.lowest = TRUE ) ) %>%
    as.data.frame( )

plotdat_violin <- plotdat_violin %>%
    mutate( tumour = factor( tumour, levels = levels( ras_sumstats_dat$tumour ) ) )
```

## Pancancer violin plot

```{r violin}
rasindex_means <- plotdat_violin %>%
    group_by( tumour ) %>%
    summarize( rasindex_mean = mean( ras_index_pc ) ) %>%
    arrange( rasindex_mean )
rasindex_means$tumour <- factor( rasindex_means$tumour, levels = rasindex_means$tumour )

quartiles <- quantile( rasindex_means$rasindex_mean, c( 1/4, 3/4 ) )
rasindex_means <- rasindex_means %>%
    mutate( iqr = rasindex_mean > quartiles[ 1 ] & rasindex_mean < quartiles[ 2 ] ) %>%
    as.data.frame( )

gg_violin <- plotdat_violin %>%
    mutate( mut_ras_type = ifelse( mut_ras_type == "amp", "wt", mut_ras_wt ) ) %>%
    ggplot( aes( tumour, ras_index_pc,
                         color = mut_ras_type ) ) +
    geom_jitter( data = subset( plotdat, !MUT.RAS ),
                size = 0.2, alpha = 0.5, width = 0.3 ) +
    geom_jitter( data = subset( plotdat, MUT.RAS ),
                 aes( reorder( tumour, ras_index_pc, mean ), ras_index_pc, color = mut_ras_type ),
                size = 1, alpha = 0.8, width = 0.3 ) +
    scale_colour_manual( values = c( my_blue, my_red, my_orange, "black" ) ) +
    coord_flip( ) +
    labs( x = "TCGA Tumour Type", y = "RAS Activity Index" ) +
    plot_formatter( ) +
    theme( legend.position = "none" ) 

gg_violin

gg_rasmut_bar <- plotdat_violin %>%
    mutate( mut_ras_type = factor( mut_ras_type, levels = c( "wt", "amp", "kras", "nras", "hras" ) ) ) %>%
    ggplot( aes( x = tumour, fill = mut_ras_type ) ) +
    geom_bar( position = "fill" ) +
    scale_y_continuous( labels = scales::percent_format( ) ) +
    scale_fill_manual( values = c( "white", "grey", my_red, my_orange, my_blue ) ) +
    coord_flip( ) +
    theme_minimal() +
    theme( axis.title.y = element_blank( ) ) +
    theme( axis.title.y = element_blank( ),
           axis.text.y = element_blank( ),
           axis.ticks.y = element_blank( ) )
   

cairo_pdf( file = file.path( adat$plot.path, "pancancer_violin_isoRAS_withbar.pdf" ),
     width = 9, height = 7 )
cowplot::plot_grid( gg_violin, gg_rasmut_bar,
                   align = "h", axis = "l", nrow = 1, rel_widths = c(0.8, 0.25 ) )
dev.off()

pdf( file = file.path( adat$plot.path, "pancancer_violin_isoRAS.pdf" ),
     width = 5, height = 7 )
gg_violin
dev.off( )
```

## RAS index distributions

Here we compare the RAS index distribution across tumour types. The violin plot below shows the distributions of the RAS index means. As observed
 they fall into two groups indicated by the bimodal distribution. We split the tumour types into high ras and normal tumour groups. The high RAS 
group contains four of the five most RAS mutated cohorts (>30%). 

```{r ras_index_dist}
d <- density( ras_sumstats_dat$mean_ras_index )

png( file = file.path( adat$plot.path, "density.png" ), height = 300 )
print( plot( d ) )
dev.off( )

max1_ind <- which.max( d$y )
max1_x <- d$x[ max1_ind ]

min1_f <- d$x > max1_x & d$x < 0.6
min1_ind <- which.min( d$y[ min1_f ] )
min1_x <- d$x[ min1_f ][ min1_ind ]

max2_f <- d$x > min1_x & d$x < 0.7
max2_ind <- which.max( d$y[ max2_f ] )
max2_x <- d$x[ max2_f ][ max2_ind ]

mean_RSI_gg <- ggplot( ras_sumstats_dat, aes( x = mean_ras_index ) ) +
    geom_density( ) +
    geom_vline( xintercept = c( max1_x, max2_x, min1_x ),
               color = c( my_red, my_red, my_orange ) ) +
    labs( x = "mean RSI / TCGA cohort", title = "" )

cairo_pdf( file = file.path( adat$plot.path, "mean_ras_index_density.pdf" ),
          width = 4, height = 4 )
print( mean_RSI_gg )
dev.off( )

mean_ras_index_violin_dat <- ras_sumstats_dat %>%
    arrange( RASmut ) %>%
    mutate( RAS_tumour_group = ifelse( mean_ras_index > 0.53, "high_ras_tumour", "tumour" ) )

high_RAS_tumour_group <- mean_ras_index_violin_dat %>%
    filter( RAS_tumour_group == "high_ras_tumour" ) %>%
    arrange( desc( RASmut ) ) %>%
    pull( tumour ) %>%
    as.character( )

mean_ras_index_violin_gg <- ggplot( mean_ras_index_violin_dat,
                                   aes( x = "tumour", y = mean_ras_index,
                                        color = RASmut, size = RASmut ) ) +
    geom_violin( ) +
    geom_point( size = 4 ) +
    scale_colour_gradient( low = "grey", high = "#cc3333" ) +
    geom_hline( yintercept = 0.53, color = my_orange, linetype = "dotted", size = 2 ) +
    geom_text_repel( data = subset( mean_ras_index_violin_dat, RASmut > 30 ),
                    aes( x = "tumour", y = mean_ras_index,
                         label = tumour, color = RASmut ),
                    inherit.aes = FALSE, nudge_x = 0.5, size = 3 ) +
    plot_formatter( ) +
    labs( x = "", y = "mean ras index" )

cairo_pdf( file = file.path( adat$plot.path, "mean_ras_index_violin.pdf" ),
          width = 2.5, height = 5 )
print( mean_ras_index_violin_gg )
dev.off( )

RAS_mutation_ratio_gg <- ggplot( ras_sumstats_dat, aes( x = RASmut/100 ) ) +
    geom_density( ) +
    geom_point( data = ras_sumstats_dat, aes( x = RASmut/100, y = 0 ) ) +
    geom_text_repel( data = subset( ras_sumstats_dat, RASmut > 30 ),
                    aes( x = RASmut/100, y = 0, label = tumour ),
                    nudge_y = 1, size = 2 ) +
    theme( panel.grid.major = elemresent_blank( ),
          panel.grid.minor = element_blank( ),
          panel.border = element_blank( ),
          panel.background = element_blank( ),
          text = element_text( size = 10 ) ) +
    labs( x = "RAS mutation ratio" )
cairo_pdf( file = file.path( adat$plot.path, "RAS_mutation_ratio_distribution.pdf" ),
          width = 4, height = 3 )
print( RAS_mutation_ratio_gg )
dev.off( )
```

## Over-represented mutations in high RAS tumour group

Here we use a hypergeometric test to determine which mutations are over-represented in the high RAS tumour group defined above.

With reference to the phyper() function

x = The number of mutated samples in high group.
m = The total number of mutated samples in the cohort.
n = The total number of wt (unmatated) samples in the cohort.
k = The umber of samples in the high group.

We remove mutations with less that 20 counts across the pancancer cohort.

```{r phyper_muts}
mut_cols <- colnames( alt_altDat )[ -1 ]
f <- colSums( as.data.frame( colData( se_pc )[, mut_cols ] ) ) > 20
f[ is.na( f ) ] <- FALSE
mut_cols <- mut_cols[ f ]
alt_phyper_pval <- lapply( mut_cols, function( mut ) {
    mut_dat <- colData( se_pc ) %>%
        as.data.frame( ) %>%
        mutate_( mutation = mut ) %>%
        mutate( high_RAS_tumours = ifelse( tumour %in% high_RAS_tumour_group, "high_RAS_tumour", "normal_RAS_tumour" ) ) %>%
        group_by( high_RAS_tumours ) %>%
        summarise( k = n(), x = sum( mutation ) )
    x <- mut_dat %>%
        filter( high_RAS_tumours == "high_RAS_tumour" ) %>%
        pull( x )
    k <- mut_dat %>%
        filter( high_RAS_tumours == "high_RAS_tumour" ) %>%
        pull( k )
    n <- mut_dat %>%
        filter( high_RAS_tumours == "normal_RAS_tumour" ) %>%
        pull( k )
    m <- mut_dat %>%
        summarise( m = sum( x ) ) %>%
        pull( m )
    data.frame( mut = mut,
               x = x, k = k, n = n, m = m,
               pval = phyper( x, m, n, k, lower.tail = FALSE ) )
} )

alt_phyper_res <- bind_rows( alt_phyper_pval ) %>%
    mutate( fdr = p.adjust( pval ) ) %>%
    arrange( pval ) %>%
    filter( fdr < 0.05 )
write.xlsx( alt_phyper_res, file = file.path( adat$analysis.path, "alt_phyper_res_high_ras_tumour_group.xlsx" ) )
```

# RAS alteration correlation with RAS index

Here we look at the correlation between mean RAS index and RAS alteration status across the cohorts. We look at RAS mutation (all three isoforms), RAS mutation and amplification (all three isoforms) and RAS pathway (any RAS pathway alteration).

## RAS mutation status

Correlation between mean RAS index and RAS mutation percentage across the cohorts.

```{r ras_mut_burden}
plotdat_rasmutburden <- lapply( se_l, function( se ) {
    ras_mut <- colData( se ) %>%
        as.data.frame( ) %>%
        mutate( MUT.RAS = MUT.KRAS | MUT.HRAS | MUT.NRAS | AMP.KRAS | AMP.HRAS ) %>%
        pull( MUT.RAS )
    data.frame( RAS_mut_burden = sum( ras_mut ) / length( ras_mut ),
                mean_ras_index = mean( se$ras_index) )
} ) %>% bind_rows( .id = "tumour" )

gg <- ggplot( plotdat_rasmutburden, aes( x = mean_ras_index, y = RAS_mut_burden ) ) +
    geom_point()
gg
```

## RAS pathway alteration status

Here we calculate the overall RAS pathway alteration burden per
sample. RAS pathway membership is defined in the pancancer cell paper. We remove cohorts with less than 100 samples.

```{r ras_pathway_alt}
ras_pathway_cols <- lapply( ras_pathway_genes, function( g ) {
    colData( se_l[[ 1 ]] ) %>%
        as.data.frame( ) %>%
        dplyr::select( contains( g, ignore.case = FALSE ) )
} ) %>% bind_cols( ) %>% colnames( )

ras_pathway_alt <- lapply( se_l,  function( se ) {
    print( se$tumour[ 1 ] )
    ras_pathway_muts <- colData( se ) %>%
        as.data.frame( ) %>%
        dplyr::select( ras_pathway_cols )
    sum( rowSums( ras_pathway_muts ) > 0 ) / nrow( ras_pathway_muts )
} ) %>% melt( ) %>%
    rename( tumour = L1 ) %>%
    rename( ras_alt_ratio = value )

cor_dat <- ras_pathway_alt %>%
    left_join( ras_sumstats_dat, by = "tumour" ) %>%
    left_join( mean_ras_index_violin_dat[, c( "tumour", "RAS_tumour_group" ) ], by = "tumour" ) %>%
    mutate( tumour = sub( "TCGA-", "", tumour ) )

cor_res <- cor.test( cor_dat$mean_ras_index, cor_dat$ras_alt_ratio, )

fit <- lm( mean_ras_index ~ ras_alt_ratio, data = cor_dat )
p1 <- predict( fit, interval = "confidence", level=0.99 )

cor_dat$outlier <- cor_dat$mean_ras_index<p1[,2] | cor_dat$mean_ras_index>p1[,3]

residuals_df <- data.frame( res_ids = residuals( fit ),
                            tumour = cor_dat$tumour ) %>%
    mutate( sd = res_ids * sd( res_ids )  ) %>%
    mutate( mad = res_ids * mad( res_ids )  ) %>%
    arrange( mad )

resids_vio_gg <- ggplot( residuals_df, aes( x = 1, y = res_ids, label = tumour ) ) +
    geom_boxplot( ) +
    geom_point( ) +
    geom_text( )
png( file = file.path( adat$plot.path, "resids_violin.png" ), width = 250 )
print( resids_vio_gg )
dev.off( )

resids_den_gg <- ggplot( residuals_df, aes( x = res_ids, label = tumour ) ) +
    geom_density( ) +
    geom_point( y = 0 ) +
    geom_text( y = 0 )
png( file = file.path( adat$plot.path, "resids_density.png" ), width = 250 )
print( resids_den_gg )
dev.off( )

gg <- ggplot( cor_dat, aes( x =  mean_ras_index, y = ras_alt_ratio,
                            label = tumour ) ) +
    geom_smooth( method = "lm", level = 0.99 ) +
    geom_point( data = cor_dat, aes(  x = mean_ras_index, y = ras_alt_ratio,
                                    color = RAS_tumour_group ) ) +
    geom_text_repel( size = 3 ) +
    scale_color_manual( values = c( my_red, my_grey ) ) +
    plot_formatter( ) +
    labs( x = "mean RAS index", y = "RAS pathway alteration ratio" ) +
    theme( legend.position = "none" )

cairo_pdf( file = file.path( adat$plot.path, "ras_burden_vs_ras_index_scatter.pdf" ),
          width = 4, height = 3 )
print( gg )
dev.off( )
```

## RAS mutant RAS index plots

Plotted here are the percentage RAS mutants per tumour cohort.

```{r rasmut_rasindex_plots,eval=FALSE}
gg <- ggplot( rasmut_dat, aes( x = RASmut ) ) +
    geom_density( ) +
    xlim( c( 0, 20 ) ) +
    labs( x = "% RAS mutant" )
gg

gg_rascount <- ggplot( plotdat, aes( x = tumour, fill = MUT.RAS ) ) +
    geom_bar() +
    scale_fill_manual( values = c( my_grey, my_red ) ) +
    plot_formatter( ) +
    theme( axis.text.x = element_text( angle = 90, hjust = 1 ) )
gg_rascount

cairo_pdf( file = file.path( adat$plot.path, "RASmut_count_barplot.pdf" ),
          width = 6, height = 3 )
gg_rascount
dev.off()
```

## RAS pathway member mutation correlation with RAS index

Here we run a wilcox test to determine if there is a significant association between RAS mutation status and ras index.

# RAS pathway alterations

Here we look at RAS index distributions across RAS pathway mutation
status. We use the RAS pathway membership definition from the
oncogenic alterations TCGA Cell paper references above. This is to
determine if any oncogenic RAS pathway driver correlates with RAS84
signature activity.

```{r alteration_alt,fig.width=8}
alt_alteration <- read.delim( file = file.path( "..", "..", "data", "oncogenicRASPath_TCGA", "oncogenic_pathway_alteration_level_TCGA.txt" ), stringsAsFactors = FALSE ) %>%
    rename( pancan_pathway = SAMPLE_BARCODE )

alt_alteration_df <- colData( se_pc ) %>%
    as.data.frame( ) %>%
    dplyr::select( tumour, "ras_index", pancan_pathway ) %>%
    left_join( alt_alteration, by = "pancan_pathway" ) %>%
    tidyr::gather( alteration, present, -ras_index, -tumour, -pancan_pathway ) %>%
    group_by( tumour, alteration ) %>% 
    mutate( prevalence = sum( present ) / n() ) %>%
    filter( prevalence > 0.05 ) %>%
    mutate( key = paste( tumour, alteration, sep = "_" ) ) %>%
    ungroup( ) %>%
    mutate( present = factor( present, levels = c( "TRUE", "FALSE" ) ) )

wilcox_fit_alteration <- alt_alteration_df %>%
    nest( -alteration, -tumour ) %>%
    mutate(
        fit = map( data, ~ wilcox.test( ras_index ~ present, data = .x ) ),
        tidied = map( fit, tidy ) ) %>% 
    unnest( tidied ) %>%
    mutate( padj = p.adjust( p.value, method = "fdr" ) ) %>%
    filter( padj < 0.01 ) %>%
    arrange( tumour ) %>%
    dplyr::select( -data, -fit, -method, -alternative ) %>%
    mutate( key = paste( tumour, alteration, sep = "_" ) ) %>%
    separate( alteration, c( "mutation_type", "gene" ), sep = "\\.", remove = FALSE ) %>%
    mutate( ras_pathway = gene %in% ras_pathway_genes ) %>%
    arrange( desc( ras_pathway ) )

DT::datatable( wilcox_fit_alteration )
```

```{r gene_alt,fig.width=8}
alt_gene <- read.delim( file = file.path( "..", "..", "data", "oncogenicRASPath_TCGA", "oncogenic_pathway_gene_level_TCGA.txt" ), stringsAsFactors = FALSE ) %>%
    rename( pancan_pathway = SAMPLE_BARCODE )

alt_gene_df <- colData( se_pc ) %>%
    as.data.frame( ) %>%
    dplyr::select( tumour, "ras_index_pc", pancan_pathway ) %>%
    left_join( alt_gene, by = "pancan_pathway" ) %>%
    tidyr::gather( gene, present, -ras_index_pc, -tumour, -pancan_pathway ) %>%
    group_by( tumour, gene ) %>% 
    mutate( prevalence = sum( present ) / n() ) %>%
    filter( prevalence > 0.05 ) %>%
    mutate( key = paste( tumour, gene, sep = "_" ) ) %>%
    ungroup( ) %>%
    mutate( present = factor( present, levels = c( "1", "0" ) ) )

wilcox_fit_gene <- alt_gene_df %>%
    nest( -gene, -tumour ) %>%
    mutate(
        fit = map( data, ~ wilcox.test( ras_index_pc ~ present, data = .x ) ),
        tidied = map( fit, tidy ) ) %>% 
    unnest( tidied ) %>%
    mutate( padj = p.adjust( p.value, method = "fdr" ) ) %>%
    filter( padj < 0.05 ) %>%
    arrange( tumour ) %>%
    dplyr::select( -data, -fit, -method, -alternative, -statistic ) %>%
    mutate( key = paste( tumour, gene, sep = "_" ) ) %>%
    mutate( ras_pathway = gene %in% ras_pathway_genes ) %>%
    arrange( desc( ras_pathway ) )

gene_lfc <- alt_gene_df %>%
    filter( key %in% wilcox_fit_gene$key ) %>%
    mutate( tumour = factor( tumour, levels = rev( rasindex_means$tumour ) ) ) %>%
    droplevels( ) %>%
    group_by( key, present ) %>%
    summarize( mean_ras_index = mean( ras_index_pc ) ) %>%
    spread( present, mean_ras_index ) %>%
    mutate( lfc = `1` - `0` ) %>%
    arrange( desc( lfc ) )
    
plotdat_gene <- alt_gene_df %>%
    filter( key %in% wilcox_fit_gene$key ) %>%
    filter( gene %in% ras_pathway_genes ) %>%
    mutate( tumour = factor( tumour, levels = rev( rasindex_means$tumour ) ) ) %>%
    droplevels( ) %>%
    arrange( tumour ) %>%
    mutate( key = factor( key, levels =  unique( key ) ) )

rec_ind <- data.frame( xmin = seq( from = 1.5, to = nlevels( plotdat$key ), by = 2 ),
                       ymin = -0.2,
                       ymax = 1.05 )
rec_ind$xmax = rec_ind$xmin + 1

dodge <- position_dodge( width = 0.6 )

alt_gene_gg <- ggplot( ) +
    geom_rect( data = rec_ind, aes( xmin = xmin, xmax = xmax,
                                    ymin = ymin, ymax = ymax ),
              fill = 'gray80', alpha = 0.5 ) +
    scale_shape_identity() +
    scale_x_discrete( labels = levels( plotdat$key ) ) +
    #geom_violin( data = plotdat, aes( x = key, y = ras_index_pc, fill = present ),
    #             position = dodge, width = 1.2 ) +
    geom_boxplot( data = plotdat, aes( x = key, y = ras_index_pc, color = present ),
                 position = dodge, width = 0.3 ) +
    scale_colour_manual( values = c( my_red, my_grey ) ) +
    plot_formatter( ) +
    theme( axis.text.x = element_text( angle = 90, hjust = 1 ),
           text = element_text( size = 8 ) )

print( alt_gene_gg )

pdf( file = file.path( adat$plot.path, "violinplot_gene.pdf" ),
    width = 7, height = 4 )
print( alt_gene_gg )
dev.off()

#gene_volc_gg <- wilcox_fit_gene %>%
#    left_join( pathway_lfc, by = "key" ) %>%
#    separate( key, c( "tumour", "gene" ), sep = "_" ) %>%
#    ggplot( aes( x = lfc, y = -log10( p.value ), color = gene ) ) +
#    geom_point( size = 2 )
#png( file = file.path( adat$plot.path, "volcanoplot_gene.png" ) )
#print( pathway_volc_gg )
#dev.off()

DT::datatable( wilcox_fit_gene )
```

```{r onco_pathway}
alt_pathway <- read.delim( file = file.path( "..", "..", "data", "oncogenicRASPath_TCGA", "oncogenic_pathway_pathway_level_TCGA.txt" ), stringsAsFactors = FALSE ) %>%
    rename( pancan_pathway = SAMPLE_BARCODE )

pathway_alt_df <- colData( se_pc ) %>%
    as.data.frame( ) %>%
    dplyr::select( tumour, "ras_index", pancan_pathway ) %>%
    left_join( alt_pathway, by = "pancan_pathway" ) %>%
    tidyr::gather( pathway, present, -ras_index, -tumour, -pancan_pathway ) %>%
    group_by( tumour, pathway ) %>% 
    mutate( prevalence = sum( present ) / n() ) %>%
    filter( prevalence > 0.05 ) %>%
    mutate( key = paste( tumour, pathway, sep = "_" ) ) %>%
    ungroup( ) %>%
    mutate( present = factor( present, levels = c( "1", "0" ) ) )

wilcox_fit_pathway <- pathway_alt_df %>%
    nest( -pathway, -tumour ) %>%
    mutate(
        fit = map( data, ~ wilcox.test( ras_index ~ present, data = .x ) ),
        tidied = map( fit, tidy ) ) %>% 
    unnest( tidied ) %>%
    mutate( padj = p.adjust( p.value, method = "fdr" ) ) %>%
    filter( padj < 0.05 ) %>%
    arrange( pathway ) %>%
    dplyr::select( -data, -fit, -method, -alternative ) %>%
    mutate( key = paste( tumour, pathway, sep = "_" ) )

pathway_lfc <- pathway_alt_df %>%
    filter( key %in% wilcox_fit_pathway$key ) %>%
    mutate( tumour = factor( tumour, levels = rev( rasindex_means$tumour ) ) ) %>%
    droplevels( ) %>%
    group_by( key, present ) %>%
    summarize( mean_ras_index = mean( ras_index ) ) %>%
    spread( present, mean_ras_index ) %>%
    mutate( lfc = `1` - `0` ) %>%
    arrange( desc( lfc ) )
    
dodge <- position_dodge( width = 1 )

plotdat <- pathway_alt_df %>%
    filter( key %in% wilcox_fit_pathway$key ) %>%
    mutate( tumour = factor( tumour, levels = rev( rasindex_means$tumour ) ) ) %>%
    droplevels( ) %>%
    arrange( tumour ) %>%
    mutate( key = factor( key, levels =  unique( key ) ) )

pathway_alt_gg <- ggplot( plotdat, aes( x = tumour, y = ras_index, color = present ) ) +
    geom_violin( position = dodge ) +
    geom_boxplot( position = dodge, width = 0.3 ) +
    facet_grid( 1 ~ pathway, space = "free", scale = "free" ) +
    scale_colour_manual( values = c( my_red, my_grey ) ) +
    theme( axis.text.x = element_text( angle = 90, hjust = 1 ) ) +
    plot_formatter( ) +
    theme( axis.text.x = element_text( angle = 90, hjust = 1 ) )

print( pathway_alt_gg )

png( file = file.path( adat$plot.path, "violinplot_pathway.png" ),
    width = 1200 )
print( pathway_alt_gg )
dev.off()

pathway_volc_gg <- wilcox_fit_pathway %>%
    left_join( pathway_lfc, by = "key" ) %>%
    separate( key, c( "tumour", "pathway" ), sep = "_" ) %>%
    ggplot( aes( x = lfc, y = -log10( p.value ), color = pathway ) ) +
    geom_point( size = 2 )
png( file = file.path( adat$plot.path, "volcanoplot_pathway.png" ) )
print( pathway_volc_gg )
dev.off()
```

```{r rasmut_rasindex}
plotdat <- colData( se_pc ) %>%
    as.data.frame( ) %>%
    dplyr::select( barcode, tumour, ras_index_pc, rasact_3, MUT.KRAS, MUT.NRAS, MUT.HRAS ) %>%
    mutate( mut_ras_type = ifelse( MUT.KRAS, "kras",
                           ifelse( MUT.NRAS, "nras",
                           ifelse( MUT.HRAS, "hras", "wt" ) ) ) ) %>%
    mutate( MUT.RAS = mut_ras_type != "wt" )
plotdat$MUT.RAS[ is.na( plotdat$MUT.RAS ) ] <- FALSE

rasmut_rasind_dat_l <- split( plotdat, plotdat$tumour )
res_l <- lapply( rasmut_rasind_dat_l, function( dat ) {
    if( sum( dat$MUT.RAS ) > 5 ) {
        wt <- wilcox.test( ras_index_pc ~ MUT.RAS, dat )
        logfc <- dat %>%
            group_by( MUT.RAS ) %>%
            summarize( mean = mean( ras_index_pc ) )
        data.frame( pval = wt$p.value,
           logFC = ( logfc$mean[ 2 ] - logfc$mean[ 1 ] ) )
    }
} )

res <- bind_rows( res_l[ !unlist( lapply( res_l, is.null ) ) ],
                 .id = "tumour" ) %>%
    mutate( fdr = p.adjust( pval ) ) %>%
    mutate( significant = fdr < 0.05 )

gg_vol <- ggplot( res, aes( x = logFC, y = -log10( pval ),
                            color = significant, label = tumour ) ) +
    geom_point( ) +
    geom_text_repel( data = subset( res, significant ) ) +
    geom_vline( xintercept = 0, color = "black", linetype = "dotted" ) +
    scale_colour_manual( values = c( my_grey, my_red ) ) +
    plot_formatter( ) +
    theme( axis.line = element_line( color = "black" ),
           legend.position = "none" ) +
    labs( color = "RAS pathway driver", x = "OS coefficient", y = "-log10( p-value )" )
gg_vol

cairo_pdf( file = file.path( adat$plot.path, "rasindex_rasmut_vol.pdf" ),
          width = 4, height = 4 )
print( gg_vol )
dev.off( )

write.xlsx( res, file = file.path( adat$analysis.path, "rasindex_rasmut_wilcox.xlsx" ) )
DT::data.table( res )

decide <- bind_rows( res_l[ !unlist( lapply( res_l, is.null ) ) ],
                    .id = "tumour" ) %>%
    mutate( fdr = p.adjust( pval ) ) %>%
    filter( fdr < 0.05 )
```

## RAS mutation RAS index boxplots

```{r rasindex_rasmut_boxplots}
gg_box_dat <- bind_rows( rasmut_rasind_dat_l ) %>%
    filter( tumour %in% decide$tumour ) %>%
    mutate( tumour = sub( "TCGA-", "", tumour ) ) %>%
    mutate( tumour = factor( tumour, levels = c( "PAAD", "HNSC", "STAD", "LUAD",
                                                "UCEC", "BRCA", "UCS", "THYM",
                                                "THCA", "TGCT" ) ) )

rec_ind <- data.frame( xmin = seq( from = 1.5, to = nlevels( gg_box_dat$tumour ), by = 2 ),
                       ymin = -0.2,
                       ymax = 1 )
rec_ind$xmax = rec_ind$xmin + 1

xlabs <- levels( gg_box_dat$tumour )
names( xlabs ) <- 1:nlevels(  gg_box_dat$tumour )

gg_box <- ggplot( ) +
    geom_rect( data = rec_ind, aes( xmin = xmin, xmax = xmax,
                                    ymin = ymin, ymax = ymax ),
              fill = 'gray80', alpha = 0.5 ) +
    scale_shape_identity() +
    scale_x_discrete( labels = xlabs ) +
    geom_boxplot( data = gg_box_dat,
                 aes( x = tumour, y = ras_index_pc, color = MUT.RAS ),
                 position = position_dodge( 0.75 ),
                 width = 0.5 ) +
    theme( axis.text.x = element_text( angle = 90, hjust = 1 ) ) +
    labs( y = "mean RAS index", x = "TCGA cohort" ) +
    scale_colour_manual( values = c( my_grey, my_red ) ) +
    coord_cartesian( ylim = c( 0, 1.1 ) ) + 
    plot_formatter( ) +
    theme( axis.text.x = element_text( angle = 90, hjust = 1 ) )

cairo_pdf( file = file.path( adat$plot.path, "rasindex_rasmut_boxplots.pdf" ),
                             width = 7, height = 5 )
print( gg_box )
dev.off( )
```

# BRAF test in SKCM & THCA

```{r braf}
plotdat <- colData( se_pc ) %>%
    as.data.frame( ) %>%
    filter( tumour %in% c( "TCGA-SKCM", "TCGA-THCA" ) ) %>%
    mutate( tumour = factor( tumour, levels = c( "TCGA-THCA", "TCGA-SKCM" ) ) ) %>%
    select( tumour, MUT.BRAF, ras_index_pc ) %>%
    mutate( MUT.BRAF = factor( MUT.BRAF ) )

rec_ind <- data.frame( xmin = seq( from = 1.5, to = nlevels( plotdat$tumour ), by = 2 ),
                       ymin = -0.2,
                       ymax = 1 )
rec_ind$xmax = rec_ind$xmin + 1

xlabs <- levels( plotdat$tumour )
names( xlabs ) <- 1:nlevels( plotdat$tumour )

gg_box <- ggplot( ) +
    geom_rect( data = rec_ind, aes( xmin = xmin, xmax = xmax,
                                    ymin = ymin, ymax = ymax ),
              fill = 'gray80', alpha = 0.5 ) +
    scale_shape_identity() +
    scale_x_discrete( labels = xlabs ) +
    geom_boxplot( data = plotdat,
                 aes( x = tumour, y = ras_index_pc, color = MUT.BRAF ),
                 position = position_dodge( 0.75 ),
                 width = 0.5 ) +
    theme( axis.text.x = element_text( angle = 90, hjust = 1 ) ) +
    labs( y = "mean RAS index", x = "TCGA cohort" ) +
    scale_colour_manual( values = c( my_grey, my_red ) ) +
    coord_cartesian( ylim = c( 0, 1.1 ) ) + 
    plot_formatter( ) +
    theme( axis.text.x = element_text( angle = 90, hjust = 1 ) )

wt_braf_thca <- wilcox.test( ras_index_pc ~ MUT.BRAF, filter( plotdat, tumour == "TCGA-THCA" ) )
wt_braf_skcm <- wilcox.test( ras_index_pc ~ MUT.BRAF, filter( plotdat, tumour == "TCGA-SKCM" ) )

cairo_pdf( file = file.path( adat$plot.path, "brafmut_rasindex.pdf" ),
         width = 4, height = 4 )
print( gg_box )
dev.off( )
```

# PIK3CA test in CESC

```{r pik3ca}
plotdat <- colData( se_pc ) %>%
    as.data.frame( ) %>%
    filter( tumour %in% c( "TCGA-CESC" ) ) %>%
    select( tumour, MUT.PIK3CA, AMP.PIK3CA, ras_index_pc ) %>%
    mutate( BOTH.PIK3CA = MUT.PIK3CA | AMP.PIK3CA ) %>%
    mutate( MUT.PIK3CA = factor( MUT.PIK3CA ) ) %>%
    mutate( BOTH.PIK3CA = factor( BOTH.PIK3CA ) )

wt_pik3ca_cesc <- list( AMP.PIK3CA = wilcox.test( ras_index_pc ~ AMP.PIK3CA, plotdat ),
                       MUT.PIK3CA = wilcox.test( ras_index_pc ~ MUT.PIK3CA, plotdat ),
                       BOTH.PIK3CA = wilcox.test( ras_index_pc ~ BOTH.PIK3CA, plotdat ) )

plotdat <- plotdat %>%
    gather( variant, present, -ras_index_pc, -tumour ) %>%
    mutate( variant = factor( variant, levels = c( "MUT.PIK3CA", "AMP.PIK3CA", "BOTH.PIK3CA" ) ) )

rec_ind <- data.frame( xmin = seq( from = 1.5, to = nlevels( plotdat$variant ), by = 2 ),
                       ymin = 0.25,
                       ymax = 1.05 )
rec_ind$xmax = rec_ind$xmin + 1

xlabs <- levels( plotdat$variant )
names( xlabs ) <- 1:nlevels( plotdat$variant )

gg_box <- ggplot( ) +
    geom_rect( data = rec_ind, aes( xmin = xmin, xmax = xmax,
                                    ymin = ymin, ymax = ymax ),
              fill = 'gray80', alpha = 0.5 ) +
    scale_shape_identity() +
    scale_x_discrete( labels = xlabs ) +
    geom_boxplot( data = plotdat,
                 aes( x = variant, y = ras_index_pc, color = present ),
                 position = position_dodge( 0.75 ),
                 width = 0.5 ) +
    theme( axis.text.x = element_text( angle = 90, hjust = 1 ) ) +
    labs( y = "mean RAS index", x = "PIK3CA variant" ) +
    scale_colour_manual( values = c( my_grey, my_red ) ) +
    plot_formatter( ) +
    theme( axis.text.x = element_text( angle = 90, hjust = 1 ) )

cairo_pdf( file = file.path( adat$plot.path, "pik3ca_mut_rasindex_TCGA-CESC.pdf" ),
         width = 4, height = 4 )
print( gg_box )
dev.off( )
```
